\chapter{Author Contributions}\label{chapter:contributions}
The codebase used for this thesis was developed by me in collaboration with Sara Visconti. I setup the initial repository and abstract classes
for the modules and pipelines as well as the data classes for the language, which was then reviewed and refactored by Sara. Both of us modified
or added to these modules and classes as needed based on experiment requirements. We also implemented specfiic modules for the experiments.

\begin{enumerate}
    \item \textbf{Phonemes}: I implemented the RandomPhonemeModule, MostCommonPhonemeModule and LanguagePhonemeModule. I also implemented the BasicAlphabetModule.
    \item \textbf{Phonotactics}: I implemented the BasicPhonotacticsModule, CustomPhonotacticsModule and SyllableBuilderModule. 
    \item \textbf{Grammar}: I implemented the BasicGrammarModule, BaselineAgglutinativeGrammarModule, and BaselineIsolatingGrammarModule. These modules built on top of Sara's Implementations of grammar features.
    \item \textbf{Vocabulary}: I implemented the FixedVocabularyModule, FromSourceVocabularyModule, ClusterSimplifyVocabularyModule, ClusterTwoLevelVocabularyModule and ApproximatingVocabularyModule. I also used the Mapping modules implemented by Sara.
    \item \textbf{Conlang}: I implemented the LLM based Source Text Translation Module.
    \item \textbf{Evaluation}: I implemented the CompressionEvaluator, DetranslationEvaluator, HTML Summary Generator and RaceCEvaluator. I used Sara's Implementation of Zipfs Law Evaluator and BertEvaluator. Sara also added the various metrics to the DetranslationEvaluator.
    \item \textbf{LLMs}: I implemented the LocalLLM and OpenAI implementation of the Remote LLM.
\end{enumerate}

We jointly developed the conceptual design and methodology. Other contributions from Sara Visconti are not reported here since they are not mentioned in this work.

